\pagebreak
\section{G10: Actionable Patterns}
\label{sec:g10-favor-actionable-patterns}

In this section, we present G10, which champions actionable, reusable implementation patterns such as code templates, example repositories, and ADR snippets over abstract principles and guidelines. G10 ensures that developers can immediately apply and verify accountability to the proposed architectural guidelines without guesswork, bridging the gap between theory and practice.

\subsection*{Conceptual Overview}
To be done during the research.

\subsection*{Objectives}
\begin{itemize}[noitemsep,topsep=2pt]
  \item \emph{Provide templates:} To be done.
  \item \emph{Offer code references:} To be done.
  \item \emph{Reduce ambiguity:} To be done.
\end{itemize}

\subsection*{Key Principles}
\begin{itemize}[noitemsep,topsep=2pt]
  \item \emph{Principle 1:} To be done.
  \item \emph{Principle 2:} To be done.
  \item \emph{Principle 3:} To be done.
  \item \emph{Principle 4:} To be done.
  \item \emph{Principle 5:} To be done.
\end{itemize}

\subsection*{Metrics \& Verification}
\begin{itemize}[noitemsep,topsep=2pt]
  \item \emph{Pattern adoption rate:} To be done.
  \item \emph{Implementation success rate:} To be done.
  \item \emph{Ambiguity reduction score:} To be done.
  \item \emph{Template reuse frequency:} To be done.
\end{itemize}

\subsection*{Documentation Guidelines}
\begin{itemize}[noitemsep,topsep=2pt]
  \item \emph{ADR Template:} To be done.
  \item \emph{Pattern library:} To be done.
  \item \emph{Change Log Practices:} To be done.
\end{itemize}

\subsection*{Tool Capabilities}
Any open‐source or proprietary software tool used to support G10 should address:
\begin{itemize}[noitemsep,topsep=2pt]
  \item \emph{Template generator:} To be done.
  \item \emph{Code snippet repository:} To be done.
  \item \emph{Pattern validation tool:} To be done.
  \item \emph{Usage analytics:} To be done.
  \item \emph{Integration with IDEs:} To be done.
\end{itemize}

\subsection*{Literature Support Commentary}
While architectural best practices are frequently cited in high-level terms, there is a dearth of evidence-based, reusable patterns for modular monoliths. Microservices literature sometimes provides migration scripts or models~\cite{abgaz2023decomposition, jin2021}, but monolith practices remain scattered in blog posts. G10 fills this void by recommending creation and dissemination of concrete patterns such as templates, code snippets and ADR examples so that teams can implement the proposed guidelines consistently and efficiently.
