\pagebreak
\section{G7: Conway's Law}
\label{sec:g7-map-ownership-to-teams}

In this section, we present G7, which focuses on mapping module ownership to individuals or small teams to promote architectural accountability. By aligning modules with clear, stable owners and documenting decision authority, G7 leverages Conway’s Law intentionally, ensuring that communication structures reinforce module boundaries rather than degrade them.

\subsection*{Conceptual Overview}
To be done during research.

\subsection*{Objectives}
\begin{itemize}[itemsep=8pt,topsep=2pt]
  \item \emph{Assign module owners:} To be done.
  \item \emph{Define team responsibilities:} To be done.
  \item \emph{Ensure accountability:} To be done.
\end{itemize}

\subsection*{Key Principles}
\begin{itemize}[itemsep=8pt,topsep=2pt]
  \item \emph{Principle 1:} To be done.
  \item \emph{Principle 2:} To be done.
  \item \emph{Principle 3:} To be done.
  \item \emph{Principle 4:} To be done.
  \item \emph{Principle 5:} To be done.
\end{itemize}

\subsection*{Metrics \& Verification}
\begin{itemize}[itemsep=8pt,topsep=2pt]
  \item \emph{Ownership clarity score:} To be done.
  \item \emph{Decision turnaround time:} To be done.
  \item \emph{Boundary modification incidents:} To be done.
  \item \emph{ADR update compliance:} To be done.
\end{itemize}

\subsection*{Documentation Guidelines}
\begin{itemize}[itemsep=8pt,topsep=2pt]
  \item \emph{ADR Template:} To be done.
  \item \emph{Team Ownership Record:} To be done.
  \item \emph{Change Log Practices:} To be done.
\end{itemize}

\subsection*{Tool Capabilities}
Any open‐source or proprietary software tool used to support G7 should address:
\begin{itemize}[itemsep=8pt,topsep=2pt]
  \item \emph{Ownership mapping dashboard:} To be done.
  \item \emph{Responsibility tracker:} To be done.
  \item \emph{Notification on boundary changes:} To be done.
  \item \emph{Access control integration:} To be done.
  \item \emph{Audit logging:} To be done.
\end{itemize}

\subsection*{Literature Support Commentary}
Although Conway’s Law is well known, its deliberate application in early-stage startups is rarely documented. Studies on microservices teams discuss ownership models and core-committers~\cite{vitharana2024challenges}, but little guidance exists for monolithic codebases. G7 bridges this gap by prescribing explicit assignment of module ownership and by using ADRs and decision logs to maintain continuity when ownership changes.
