\pagebreak
\section{G8: DevOps Maturity}
\label{sec:g8-match-complexity-devops-maturity}

In this section, we present G8, which emphasizes adjusting architectural complexity to match DevOps maturity, avoiding premature fragmentation. By assessing automation levels, deployment processes, and observability capabilities, G8 helps teams choose the right degree of modularity so that architecture aligns with infrastructure readiness and minimizes operational risk.

\subsection*{Conceptual Overview}
To be done during research.

\subsection*{Objectives}
\begin{itemize}[itemsep=8pt,topsep=2pt]
  \item \emph{Assess DevOps maturity:} To be done.
  \item \emph{Scale architecture accordingly:} To be done.
  \item \emph{Prevent premature fragmentation:} To be done.
\end{itemize}

\subsection*{Key Principles}
\begin{itemize}[itemsep=8pt,topsep=2pt]
  \item \emph{Principle 1:} To be done.
  \item \emph{Principle 2:} To be done.
  \item \emph{Principle 3:} To be done.
  \item \emph{Principle 4:} To be done.
  \item \emph{Principle 5:} To be done.
\end{itemize}

\subsection*{Metrics \& Verification}
\begin{itemize}[itemsep=8pt,topsep=2pt]
  \item \emph{DevOps readiness score:} To be done.
  \item \emph{Operational failure rate:} To be done.
  \item \emph{Deployment frequency:} To be done.
  \item \emph{Fragmentation incidents:} To be done.
\end{itemize}

\subsection*{Documentation Guidelines}
\begin{itemize}[itemsep=8pt,topsep=2pt]
  \item \emph{ADR Template:} To be done.
  \item \emph{Maturity assessment record:} To be done.
  \item \emph{Change Log Practices:} To be done.
\end{itemize}

\subsection*{Tool Capabilities}
Any open‐source or proprietary software tool used to support G8 should address:
\begin{itemize}[itemsep=8pt,topsep=2pt]
  \item \emph{Maturity assessment dashboard:} To be done.
  \item \emph{Architecture complexity analyzer:} To be done.
  \item \emph{DevOps pipeline integration:} To be done.
  \item \emph{Fragmentation warning system:} To be done.
  \item \emph{Operational readiness checks:} To be done.
\end{itemize}

\subsection*{Literature Support Commentary}
Most architecture literature assumes a baseline level of DevOps maturity (e.g., CI/CD, logging and orchestration) without acknowledging that early-stage teams may lack such infrastructure. Practitioner accounts (e.g., Segment and Shopify) note these constraints, but academic work remains silent. G8 addresses this discrepancy by making DevOps maturity an explicit factor in architectural decisions, preventing over-engineered modularity before a team can support it.
