\pagebreak
\section{G9: Onboarding}
\label{sec:g9-treat-onboarding-first-class}

In this section, we present G9, which treats developer onboarding as a first-class architectural concern. By maintaining a unified codebase, co-locating module documentation, and providing standardized test environments, G9 accelerates new engineers’ ramp-up and preserves productivity, ensuring that growth in headcount does not compromise code quality or velocity.

\subsection*{Conceptual Overview}
To be done during research.

\subsection*{Objectives}
\begin{itemize}[noitemsep,topsep=2pt]
  \item \emph{Unified codebase:} To be done.
  \item \emph{Module documentation:} To be done.
  \item \emph{Standardized test environment:} To be done.
\end{itemize}

\subsection*{Key Principles}
\begin{itemize}[noitemsep,topsep=2pt]
  \item \emph{Principle 1:} To be done.
  \item \emph{Principle 2:} To be done.
  \item \emph{Principle 3:} To be done.
  \item \emph{Principle 4:} To be done.
  \item \emph{Principle 5:} To be done.
\end{itemize}

\subsection*{Metrics \& Verification}
\begin{itemize}[noitemsep,topsep=2pt]
  \item \emph{Onboarding time:} To be done.
  \item \emph{Documentation coverage:} To be done.
  \item \emph{First‐time test success rate:} To be done.
  \item \emph{Knowledge handoff incidents:} To be done.
\end{itemize}

\subsection*{Documentation Guidelines}
\begin{itemize}[noitemsep,topsep=2pt]
  \item \emph{ADR Template:} To be done.
  \item \emph{Onboarding checklist:} To be done.
  \item \emph{Change Log Practices:} To be done.
\end{itemize}

\subsection*{Tool Capabilities}
Any open‐source or proprietary software tool used to support G9 should address:
\begin{itemize}[noitemsep,topsep=2pt]
  \item \emph{Documentation generator:} To be done.
  \item \emph{Test environment provisioning:} To be done.
  \item \emph{Onboarding tracker:} To be done.
  \item \emph{Unified repository manager:} To be done.
  \item \emph{Mentorship assignment system:} To be done.
\end{itemize}

\subsection*{Literature Support Commentary}
Few studies, whether on monoliths or microservices, address onboarding as an architectural issue~\cite{montesi2021sliceable}. Without clear module docs, separate repositories, or scriptable environments, new hires struggle to learn the system, slowing development. G9 remedies this by incorporating onboarding artifacts like README files, test fixtures, and guided tutorials directly into the modular structure, thereby reducing cognitive overhead for newcomers.
